---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# cytometree

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/cytometree)](https://cran.r-project.org/package=cytometree)
[![Travis-CI Build Status](https://travis-ci.org/chariff/Cytometree.svg?branch=master)](https://travis-ci.org/chariff/cytometree)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/chariff/cytometree?branch=master&svg=true)](https://ci.appveyor.com/project/chariff/cytometree)
[![codecov.io](https://codecov.io/github/chariff/Cytometree/coverage.svg?branch=master)](https://codecov.io/github/chariff/Cytometree?branch=master)
[![Downloads](https://cranlogs.r-pkg.org/badges/cytometree?color=blue)](https://www.r-pkg.org/pkg/cytometree)


## Overview

`cytometree` is a package which performs automatic gating and annotation of flow-cytometry
data. On top of the [CRAN help files](https://cran.r-project.org/web/packages/cytometree/cytometree.pdf), 
we also provide a [vignette](https://cran.r-project.org/web/packages/cytometree/vignettes/autogating_cytometree.html)
illustrating the functionnalities of `cytometree`.

The following article explains what `cytometree` is about:

> D Commenges, C Alkhassim, R Gottardo, BP Hejblum, R Thiébaut, cytometree: a binary tree algorithm for automatic gating in cytometry analysis, 2017, *in revision*.


## Installation

The easiest way to get `cytometree` is to install it from [CRAN](https://cran.r-project.org/package=cytometree):
```{r CRAN-install, eval=FALSE}
install.packages("cytometree")
```

Or to get the development version from [GitHub](https://github.com/chariff/cytometree):
```{r GitHub-install, eval=FALSE}
#install.packages("devtools")
devtools::install_github("chariff/cytometree")
```

-- Chariff Alkhassim & Boris Hejblum


