---
title: "User guide for automatic gating with cytometree"
author: "Chariff Alkhassim, Boris Hejblum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Introduction to `cytometree`

`cytometree` is a package that implements a binary tree algorithm for the analysis of cytometry data.

Its core algorithm is based on the construction of a binary tree, whose nodes represents 
cell subpopulations.

## Binary tree construction

1. At each node, observed cells (or "events") and markers are modeled by both a normal distribution (so *uni-modal*), and a mixture of 2 normal distributions (so *bi-modal*).

2. Splitting of the events at each node is done according to a normalized difference of AIC between 
the two distributional fit (unimodal or bimodal), allowing to pick the marker that best splits those data.

3. When AIC differences are not significant anymore, the tree and the cells have been automatically gated (i.e. partitioned). 

## Post-hoc annotation

Given the unsupervised nature of the binary tree, some of the available markers may not be used to find the different cell populations present in a given sample. To recover a complete annotation, we defined, as a post processing procedure, an annotation method which allows the user to distinguish two (or three) expression levels per marker.


# Example of an analysis with `cytometree`

In this example, we will use a diffuse large B-cell lymphoma dataset (from the [flowCAP-I challenge](http://flowcap.flowsite.org/), with only 3 markers. First, we need to load the package `cytometree` and we can have a look at the data:

```{r, message=FALSE}
library(cytometree)
dim(DLBCL)
head(DLBCL)
```
We have 3 markers measured `FL1`, `FL2`, `FL4` as well as the `label` obtained from manual gating, and `r nrow(DLBCL)`.

```{r}
# Cell events.
cellevents <- DLBCL[,c("FL1", "FL2", "FL4")]
# Manual gating reference from FlowCAP-I
manual_labels <- DLBCL[,"label"]
```


```{r, eval=FALSE}
# Build the binary tree.
Tree <- CytomeTree(cellevents, minleaf = 1, t=.1)

# Retreive the resulting partition of the set N.
Tree_Partition <- Tree$labels
# Plot node distributions.
par(mfrow=c(1,2))
plot_nodes(Tree)
# Choose a node to plot.
plot_nodes(Tree,"FL4.1")
# Plot a graph of the tree.
par(mfrow=c(1,1))
plot_graph(Tree,edge.arrow.size=.3, Vcex =.5, vertex.size = 30)
# Run the annotation algorithm.
Annot <- Annotation(Tree,plot=FALSE)
Annot$combinations
# Compare to the annotation gotten from the tree.
Tree$annotation
# Example of seeked phenotypes.
# Variable in which seeked phenotypes can be entered in the form
# of matrices.
phenotypes <- list()
## Seeked phenotypes.
# FL1- FL2+ FL4-.
phenotypes[[1]] <- rbind(c("FL1", 0), c("FL2", 1), c("FL4", 0))
# FL1+ FL2- FL4+.
phenotypes[[2]] <- rbind(c("FL1", 1), c("FL2", 0), c("FL4", 1))
# FL1+ FL2+ FL4+.
phenotypes[[3]] <- rbind(c("FL1", 1), c("FL2", 1), c("FL4", 1))
# Retreive cell populations found using Annotation.
PhenoInfos <- RetrievePops(Annot, phenotypes)
PhenoInfos$phenotypesinfo
# F-measure ignoring cells labeled 0 as in FlowCAP-I.
# Use FmeasureC() in any other case.
FmeasureC_no0(ref=manual_labels, pred=Tree_Partition)
# Scatterplots.
library(ggplot2)
# Ignoring cells labeled 0 as in FlowCAP-I.
rm_zeros <- which(!manual_labels)
# Building the data frame to scatter plot the data.
FL1 <- cellevents[-c(rm_zeros),"FL1"]
FL2 <- cellevents[-c(rm_zeros),"FL2"]
FL4 <- cellevents[-c(rm_zeros),"FL4"]
n <- length(FL1)
Labels <- c(manual_labels[-c(rm_zeros)]%%2+1, Tree_Partition[-c(rm_zeros)])
Labels <- as.factor(Labels)
method <- as.factor(c(rep("FlowCap-I",n),rep("CytomeTree",n)))
scatter_df <- data.frame("FL2"=FL2,"FL4"=FL4,"labels"=Labels,"method"=method)

p <- ggplot2::ggplot(scatter_df,  ggplot2::aes_string(x = "FL2", y="FL4",colour="labels"))+
 ggplot2::geom_point(alpha = 1,cex = 1)+
 ggplot2::scale_colour_manual(values = c("green","red","blue"))+
 ggplot2::facet_wrap(~ method)+
 ggplot2::theme_bw()+
 ggplot2::theme(legend.position="bottom")
p
```


Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
