---
title: "`cytoftree`: extension of `cytometree` to analyze mass cytometry data"
author: "Anthony Devaux, Boris Hejblum"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{`cytoftree`: extension of `cytometree` to analyze mass cytometry data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitrsetup, include=FALSE}
knitr::opts_chunk$set(tidy = TRUE)
knitr::knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar = c(0, 0, 0, 0))  # no margin
})
```

# Introduction to `cytoftree`

`cytoftree` is an extension to `cytometree` function to analyze mass cytometry data. These data are specific due to a high number of zero and the high number of markers (up to 100 potentially). `cytoftree` is based on `cytometree`'s algorithm which is the construction of binary tree, whose nodes represents cell subpopulations, and slighly modified to take into account the specification of mass cytometry data.

## Data transformation

According to the litterature, mass cytometry data must be transform to get best partitions. We propose differents transformations: `asinh` (as default), `biexp`, `log10` or `none` (without transformation).

## Binary tree construction

1. At each node, for each marker, the cells with zero values are temporarily set aside from the other cells.

2. The remaining observed cells (or "events") and markers are modeled by both a normal distribution (so *unimodal*), and a mixture of 2 normal distributions (so *bimodal*).

3. If the AIC normalized differences $D$ are significant, the cells are split into 2 groups according to the bimodal distribution. Cells with low values are annoted `-` (no marker) while cells with high values are annoted `+` (with marker). The cells with zero values are also annoted `-` (no marker).

4. The binary tree is constructed until the cells can no longer be split into 2 groups.

## Post-hoc annotation

Given the unsupervised nature of the binary tree, some of the available markers may not be used to find the different cell populations present in a given sample. To recover a complete annotation, we defined, as a post processing procedure, an annotation method which allows the user to distinguish two (or three) expression levels per marker.

# Influenza vaccine response dataset analysis with `cytoftree`

In this example, we will use an influenza vaccine response dataset (from [ImmuneSpace](https://www.immunespace.org/project/Studies/SDY478/begin.view?)), with 39 markers. To speed-up the computation, we sampled 10 000 cells from this dataset.

## Data preparation

First, we can look the structure and the markers of the data.

```{r, message = FALSE, warning = FALSE}
library(cytometree)
data(IMdata)
dim(IMdata)
colnames(IMdata)
```

Then, we also check the proportion of zero for each marker, particularity of mass cytometry data.

```{r}
zero_proportion <- apply(IMdata[,-c(1,2)], 
                         MARGIN = 2, 
                         FUN = function(x){round(prop.table(table(x==0))["TRUE"]*100,2)})
zero_proportion
```

## CytofTree function

According to the available markers, a gating strategy may be considered. In this example, we have a gating strategy to conserve only viable cells by splitting on the following markers : `DNA1`, `DNA2`, `Cell_length`, `Bead` and `Dead`. This way, we can be as close as possible to manual gating. To do this, we have to force the markers with the `force_first_marker` option (semi-supervised gating).

Then, to improve the performance of automating gating, we decided to transform data with `asinh` transformation (default transformation). Then, we have to choose which markers should be tranformed using `num_col` argument. The columns `Time` et `Cell_length` are not mass cytometry measure and shouldn't be transformed.

```{r}
num_col <- c(3:ncol(IMdata))

tree <- CytofTree(M = IMdata,
                  minleaf = 1,
                  t = 0.1,
                  verbose = FALSE,
                  force_first_markers = c("(Ir191)Dd_DNA1",
                                          "(Ir193)Dd_DNA2",
                                          "Cell_length",
                                          "(Ce140)Dd_Bead",
                                          "(In115)Dd_Dead"),
                  transformation = "asinh",
                  num_col = num_col)

max(tree$labels)
```

## High dimensional issues

Due to the high number of markers, `cytoftree` provides high number of subpopulations. `minleaf` value for the minimum of cells by subpopulation and `t` threshold for the depth of the binary tree can be modified to get more or less subpopulations. The `plot_graph` function provides a look on the binary tree, but should be unreadable due to the high number of subpopulations. 

## Annotation function

The `annotation` function allows to recover the incomplete annotation on subpopulations. `combinations` option provides the complete annotation on each subpopulation.

```{r}
annot <- Annotation(tree, plot = FALSE, K2markers = colnames(IMdata))
annot$combinations[1:5,]
```

Due to the high number of subpopulations, it's recommanded to use `RetrievePops` function which provide informations for particular subpopulations.

## `RetrievePops` : providing informations for particular subpopulations

`RetrievePops` provides several informations on specific subpopulations, in particular the proportions and the subpopulations merged.

```{r}
phenotypes <- list()
phenotypes[["CD4+"]] <- rbind(c("(Ir191)Dd_DNA1", 1), c("(Ir193)Dd_DNA2", 1), 
                              c("Cell_length", 0), c("(Ce140)Dd_Bead", 0), 
                              c("(In115)Dd_Dead", 0), c("(Sm154)Dd_CD14", 0), 
                              c("(Er166)Dd_CD33", 0), c("(Nd150)Dd_CD3", 1), 
                              c("(Nd143)Dd_CD4", 1))

phenotypes[["CD8+"]] <- rbind(c("(Ir191)Dd_DNA1", 1), c("(Ir193)Dd_DNA2", 1), 
                              c("Cell_length", 0), c("(Ce140)Dd_Bead", 0), 
                              c("(In115)Dd_Dead", 0), c("(Sm154)Dd_CD14", 0), 
                              c("(Er166)Dd_CD33", 0), c("(Nd150)Dd_CD3", 1), 
                              c("(Nd144)Dd_CD8", 1))

pheno_result <- RetrievePops(annot, phenotypes = phenotypes)

# CD4+
pheno_result$phenotypesinfo[[1]]

# CD8+
pheno_result$phenotypesinfo[[2]]
```

## Proportions comparison between manual and automatic gating

We can compare proportions providing by automatic gating (`cytoftree`) and manual gating for the selected subpopulations.

```{r, echo = FALSE}
automating <- c(pheno_result$phenotypesinfo[[1]]$proportion,
                pheno_result$phenotypesinfo[[2]]$proportion)
manual <- c(0.1824389, 0.06523925)
resu <- rbind(manual, automating)
rownames(resu) <- c("Manual Gating", "Automating Gating")
colnames(resu) <- c("CD4+", "CD8+")
knitr::kable(resu, digits = 3)
```

`cytoftree` provides good results, close to the proportions getting by manual gating.