---
title: "`cytoftree`: extension of `cytometree` to analyze mass cytometry data"
author: "Anthony Devaux, Boris Hejblum"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{`cytoftree`: extension of `cytometree` to analyze mass cytometry data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitrsetup, include=FALSE}
knitr::opts_chunk$set(tidy = TRUE)
knitr::knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar = c(0, 0, 0, 0))  # no margin
})
```

# Introduction to `cytoftree`

`cytoftree` is an extension to `cytometree` function to analyze mass cytometry data. These data are specific due to a high number of zero and the high number of markers (up to 100 potentially). `cytoftree` is based on `cytometree`'s algorithm which is the construction of binary tree, whose nodes represents cell subpopulations, and slighly modified to take into account the specification of mass cytometry data.

## Data transformation

According to the litterature, mass cytometry data must be transform to get best partitions. We propose differents transformations: `asinh` (as default), `none` (without transformation).

## Binary tree construction

1. At each node, for each marker, the cells with zero values are set apart from the other cells.

2. The remaining observed cells (or "events") and markers are modeled by both a normal distribution (so *unimodal*), and a mixture of 2 normal distributions (so *bimodal*).

3. If the AIC normalized differences $D$ are significant, the cells are split into 2 groups according to the bimodal distribution. Cells with low values are annoted `-` (no marker) while cells with high values are annoted `+` (with marker). The cells with zero values are also annoted `-` (no marker).

4. The binary tree is constructed until the cells can no longer be split into 2 groups.

## Post-hoc annotation

Given the unsupervised nature of the binary tree, some of the available markers may not be used to find the different cell populations present in a given sample. To recover a complete annotation, we defined, as a post processing procedure, an annotation method which allows the user to distinguish two (or three) expression levels per marker.

# Influenza vaccine response dataset analysis with `cytoftree`

In this example, we will use an influenza vaccine response dataset (from [ImmuneSpace](https://www.immunespace.org/project/Studies/SDY478/begin.view?)), with 39 markers. To speed-up the computation of this example, we sampled 10 000 cells from this dataset.

## Data preparation

First, we can look the structure and the markers of the data

```{r, message = FALSE, warning = FALSE}
library(cytometree)
data(IMdata)
dim(IMdata)
colnames(IMdata)
```

Then, we also check the proportion of zero for each marker, particularity of mass cytometry data.

```{r}
zero_proportion <- apply(IMdata[,-c(1,2)], 
                         MARGIN = 2, 
                         FUN = function(x){round(prop.table(table(x==0))["TRUE"]*100,2)})
zero_proportion
```

## CytofTree function

To improve the performance of automating gating, we decided to transform data with `asinh` transformation. Then, we have to choose which markers must be tranform using `num_col` argument. The columns `Time` et `Cell_length` are not mass cytometry measure and mustn't be transform.

### Without `force_first_marker` option

Without `force_first_marker` option, we let `cytoftree` choose the best marker to split the cells on each node.

```{r}
num_col <- c(3:ncol(IMdata))

tree <- CytofTree(M = IMdata,
                  minleaf = 1,
                  t = 0.1,
                  verbose = FALSE,
                  force_first_markers = NULL,
                  transformation = "asinh",
                  num_col = num_col)

max(tree$labels)
```

### With `force_first_marker` option

While with `force_first_marker` option, `cytoftree` splits on particular markers. Specially in this example, we force `cytoftree` to split on the markers `DNA1`, `DNA2`, `Cell_length`, `Bead` and `Dead` to be the closest to the manual gating.

```{r}
tree_force <- CytofTree(M = IMdata,
                        minleaf = 1,
                        t = 0.1,
                        verbose = FALSE,
                        force_first_markers = c("(Ir191)Dd_DNA1",
                                                "(Ir193)Dd_DNA2",
                                                "Cell_length",
                                                "(Ce140)Dd_Bead",
                                                "(In115)Dd_Dead"),
                        transformation = "asinh",
                        num_col = num_col)

max(tree_force$labels)
```

## High dimensional issues

Due to the high number of markers, `cytoftree` provides high number of subpopulations, but can be decrease modifying `minleaf` value for the minimum of cells by subpopulation and `t` threshold for the depth of the binary tree. 

The `plot_graph` function provides a look on the binary tree, but should be unreadable due to the high number of subpopulations. Nevertheless, we can take a look on the distribution on a particular node with `plot_nodes` function.

```{r}
plot_nodes(tree, nodes = "(Dy164)Dd_CD20.2")
```

